Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves

### 2025-07-23 ms/ms.qmd
loo table 2

---
title: Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: https://www.zotero.org/styles/science
---

@@

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019]. When LMA is included in the model, the effect of growth light intensity on \aax{} is no longer significant based on model comparison (Table S8) and LMA is weakly positively correlated with \aax{} ([Fig. @fig-lma_aa]). In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table S8). Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted. Hence, amphistomy may, expectedly, be particularly beneficial for sun leaves experiencing intermittant shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

@@

>>>>

---
title: Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: science.csl
---

@@

### MARKER - where is evidence for relationship between PAI and LMA, AA?

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides modest support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019]. While LMA is weakly, albeit significantly, associated with individual-level \aax{} ([Fig. @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1). In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table/Figure). Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted. Hence, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

@@

### MARKER - make other LOOIC table

```{r, aa_loo2, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-aa_loo2
#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, leaf mass per area (LMA), and population. The model with the lowest LOOIC is considered the best fit to the data. The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC). Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed effects, random effects, or distributional effect on the residual variance. We test for fixed interaction effects between measurement and growth light intensity and LMA. We also test for random interaction effects between measurement and growth light intensity. All models included phylogenetically structured random effects of population.'

df1 = aa_loo_table2 |>
  mutate(
    fixed_meas = str_detect(fixed, "light_intensity"),
    fixed_growth = str_detect(fixed, "light_treatment"),
    fixed_lma = str_detect(fixed, "mi\\(llma\\)"),
    fixed_intrxn123 = str_detect(fixed, "light_treatment \\* light_intensity \\* mi\\(llma\\)"),
    fixed_intrxn12 = str_detect(fixed, "light_treatment \\* light_intensity") | fixed_intrxn123,
    fixed_intrxn13 = str_detect(fixed, "light_treatment \\* mi\\(llma\\)") | fixed_intrxn123,
    fixed_intrxn23 = str_detect(fixed, "light_intensity \\* mi\\(llma\\)") | fixed_intrxn123,
    # random_acc = TRUE,
    random_meas = str_detect(random, "light_intensity"),
    random_growth = str_detect(random, "light_treatment"),
    random_intrxn = str_detect(random, "\\*"),
    sigma_meas = str_detect(sigma, "light_intensity"),
    sigma_growth = str_detect(sigma, "light_treatment"),
  ) |>
  dplyr::select(
    LOOIC = looic,
    `$\\Delta$LOOIC` = delta_looic,
    starts_with("fixed_"),
    starts_with("random_"),
    starts_with("sigma_")
  ) |>
  mutate(across(where(is.logical), ~ ifelse(.x, "$\\checkmark$", "")))

c2 = case_when(
  str_detect(c1, "_meas") ~ "Meas.",
  str_detect(c1, "_growth") ~ "Growth",
  str_detect(c1, "_lma") ~ "LMA",
  str_detect(c1, "_intrxn12") ~ "Meas. x Growth",
  str_detect(c1, "_intrxn13") ~ "Meas. x LMA",
  str_detect(c1, "_intrxn23") ~ "Growth x LMA",
  str_detect(c1, "_intrxn123") ~ "Meas. x Growth x LMA",
  str_detect(c1, "_acc") ~ "Population",
  .default = c1
)

### 2025-07-23 ms/ms.tex
loo table 2

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, argues against the acclimatory hypothesis, and
provides only weak support for the constitutive hypothesis. Plastic
changes in leaf thickness and/or packing density, summarized by the bulk
leaf mass per area (LMA), likely mediate the effect of growth light
intensity on \aax{}. LMA increased in sun grown plants in all
populations by an average of 119\% {[}95\% CI: 47 to 240\%{]}, similar
to plastic responses in many species (\emph{48}). When LMA is included
in the model, the effect of growth light intensity on \aax{} is no
longer significant based on model comparison (Table S8) and LMA is
weakly positively correlated with \aax{} (Fig.~\ref{fig-lma_aa}). In
contrast, LMA did not mediate an indirect effect of native PAI on \aax{}
(Table S8). Many anatomical traits underlie LMA (\emph{53}) and future
research will be needed identify which particular traits, such as leaf
thickness or mesophyll porosity, are responsible for mediating \aax{}.
The fact that the \aax{} of sun plants was greater under low measurement
light intensity supports a long-standing hypothesis that resistance to
CO\(_2\) diffusion is greater in the upper than lower portions of the
leaf interior (\emph{54}). At low light intensity, the bulk properties
of the leaf are weighed toward the upper palisade where most light is
intercepted. Hence, amphistomy may, expectedly, be particularly
beneficial for sun leaves experiencing intermittant shade or cloud
cover. Our study is limited in testing this because we could not
directly measure the stomatal conductance ratio and intercellular
resistance on each surface. Future experiments measuring \aax{} with a
dual sided chamber (\emph{31}, \emph{33}) can overcome these
limitations.

@@

>>>>

\subsubsection{MARKER - where is evidence for relationship between PAI
and LMA,
AA?}\label{marker---where-is-evidence-for-relationship-between-pai-and-lma-aa}

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, argues against the acclimatory hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 119\% {[}95\% CI: 47 to 240\%{]}, similar to plastic
responses in many species (\emph{48}). While LMA is weakly, albeit
significantly, associated with individual-level \aax{}
(Fig.~\ref{fig-lma_aa}), the effect of growth light intensity on \aax{}
is still predictive based on model comparison (Table~\ref{tbl-aa_loo1}).
In contrast, LMA did not mediate an indirect effect of native PAI on
\aax{} (Table/Figure). Many anatomical traits underlie LMA (\emph{53})
and future research will be needed identify which particular traits,
such as leaf thickness or mesophyll porosity, are responsible for
mediating \aax{}. The fact that the \aax{} of sun plants was greater
under low measurement light intensity supports a long-standing
hypothesis that resistance to CO\(_2\) diffusion is greater in the upper
than lower portions of the leaf interior (\emph{54}). At low light
intensity, the bulk properties of the leaf are weighed toward the upper
palisade where most light is intercepted. Hence, amphistomy may,
unexpectedly, be particularly beneficial for sun leaves experiencing
intermittent shade or cloud cover. Our study is limited in testing this
because we could not directly measure the stomatal conductance ratio and
intercellular resistance on each surface. Future experiments measuring
\aax{} with a dual sided chamber (\emph{31}, \emph{33}) can overcome
these limitations.

@@

\subsubsection{MARKER - make other LOOIC
table}\label{marker---make-other-looic-table}

\caption{\label{tbl-aa_loo2}Leave-one-out cross-validation information
criterion (LOOIC) for models of \aax{} as a function of light intensity,
light treatment, leaf mass per area (LMA), and population. The model
with the lowest LOOIC is considered the best fit to the data. The LOOIC
column lists the LOOIC value for each model; the \(\Delta\)LOOIC column
lists the difference between the LOOIC of the corresponding model and
the best model (lowest LOOIC). Checkmarks in each row indicate whether
the corresponding model included affects of measurement light intensity
(``Meas.''), growth light intensity (``Growth''), or LMA as fixed
effects, random effects, or distributional effect on the residual
variance. We test for fixed interaction effects between measurement and
growth light intensity and LMA. We also test for random interaction
effects between measurement and growth light intensity. All models
included phylogenetically structured random effects of population.}
